---
title: "CDC Data Exercise"
---

This data analysis was done using the *Provisional COVID-19 Deaths by Race and Hispanic Origin, and Age* dataset provided by the NCHS/DVS, which contains data on deaths involving COVID-19, pneumonia, and influenza, categorized by race/age/jurisdiction.

There are 13 columns listed:

1.  Date as of - date of analysis

2.  Start date - first date of data period

3.  End date - last date of data period

4.  State - jurisdiction of occurrence

5.  Age group - age group

6.  Race and Hispanic Origin Group - Race or Hispanic origin group

7.  COVID-19 deaths - deaths involving COVID-19 (ICD-code U07.1)

8.  Total deaths - deaths from all causes of death

9.  Pneumonia deaths - Pneumonia deaths (ICD-10 codes J12.0-J18.9)

10. Pneumonia and COVID-19 deaths - Deaths with Pneumonia and COVID-19

11. Influenza deaths - Infleunza deaths (ICD-10 codes J09-J11)

12. Pneumonia, Influenza, or COVID-19 deaths - deaths with any of aforementioned codes.

13. Footnote- suppressed counts (1-9)

## Package loading and install

First we need to upload and/or install packages for cleaning our data
```{r}
# install packages if not already installed (run once in the console)
# load required packages
# tidyverse added for later steps with data processing
library("dslabs")
library("tidyverse")
library("readr")
library("stringr")
library("here")
library("janitor") 
```

```{r}
# read in the data
cdcdata <- read_csv(here("cdcdata-exercise", "Provisional_COVID-19_Deaths_by_Race_and_Hispanic_Origin,_and_Age_20260211.csv"))
```

## Show dataset structure
Now we should take a look at what the data actually looks like.
```{r}
# get an overview of data structure
# I prefer str() for checking out the data, but I also like glimpse() for very rote summaries where you can determine characteristics.We can run all of them and see what we get, and then just choose whatever we'd like.
str(cdcdata)
glimpse(cdcdata)
```

## Data processing

### Cleaning up dataset

```{r}
library(dplyr)
library(readr)

# clean data
cdc_clean <- cdcdata %>%
  # remove columns not needed for our analysis
  select(-`Data as of`, -`Start Date`, -Footnote) %>%
  # rename columns-- I do this as the original dataset would require backticks for uppercase and spaced column names.
  rename(
    end_date = `End Date`,
    state = State,
    age_group = `Age group`,
    race_origin = `Race and Hispanic Origin Group`,
    covid_deaths = `COVID-19 Deaths`,
    total_deaths = `Total Deaths`,
    pneumonia_deaths = `Pneumonia Deaths`,
    pneumonia_covid_deaths = `Pneumonia and COVID-19 Deaths`,
    influenza_deaths = `Influenza Deaths`,
    pneumonia_influenza_covid_deaths = `Pneumonia, Influenza, or COVID-19 Deaths`
  ) %>%
# standardize formats and clean spacing. `mdy()` converts the end_date column to a date format, while `str_squish()` and `str_trim()` remove extra spaces from the state. I noticed that some state values were returning as `NA` because of extra spaces. So look out for that!
  mutate(
    end_date = mdy(end_date),
    state = str_squish(str_trim(state)),
    age_group = str_squish(str_trim(age_group)),
    race_origin = str_squish(str_trim(race_origin))
  )
# also saw that NYC was separated from New York, so I combined those two together to avoid confusion and make sure all data from New York is in one place.
cdc_clean <- cdc_clean %>%
  mutate(state = ifelse(state %in% c("New York", "New York City"), "New York", state))
# now we can double check that names are correct and data is in the correct format.
names(cdc_clean)
glimpse(cdc_clean)
```

## Analysis and Plotting

We can now start to analyze the data and make some plots. For example, we can look at the relationship between COVID-19 deaths and total deaths across different age groups or states. 

We can also explore how pneumonia deaths relate to COVID-19 deaths, or how influenza deaths compare across different race and Hispanic origin groups.

```{r}
library(dplyr)
library(ggplot2)
library(scales)

state_totals <- cdc_clean %>%
  filter(state != "United States") %>%
  group_by(state) %>%
  summarise(total_deaths = sum(total_deaths, na.rm = TRUE), .groups = "drop")
# this gives us the top 10 states by total deaths, which we can plot in a horizontal bar chart.
top10 <- state_totals %>%
  arrange(desc(total_deaths)) %>%
  slice_head(n = 10)
# horizontal bar chart: states on y-axis, total_deaths on x-axis 
ggplot(top10, aes(x = reorder(state, total_deaths), y = total_deaths, fill = state)) +
  geom_col() +
  geom_text(
    aes(label = comma(total_deaths)),
    hjust = 1.1,           # move text slightly inside bar
    color = "white",
    fontface = "bold",
    size = 3.5
  ) +
  coord_flip() + # flip coordinates for horizontal bars
  scale_fill_hue() +
  scale_y_continuous(
    breaks = seq(0, max(top10$total_deaths, na.rm = TRUE), by = 250000),
    labels = comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Top 10 States by Total Deaths",
    x = "State",
    y = "Total Deaths"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

As seen above, California has the highest total deaths, followed by Texas and Florida. The top 10 states account for a significant portion of the total deaths in the dataset. Now we can look at the relationship between COVID-19 deaths and total deaths across different states and variables.
```{r}
library(dplyr)
library(ggplot2)
library(scales)

# top 10 states by covid deaths (excluding US aggregate; merged NYC into NY)
covid_top10 <- cdc_clean %>%
  filter(state != "United States") %>%
  mutate(state = ifelse(state == "New York City", "New York", state)) %>%
  group_by(state) %>%
  summarise(covid_deaths = sum(covid_deaths, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(covid_deaths)) %>%
  slice_head(n = 10)
ggplot(
  covid_top10,
  aes(y = reorder(state, covid_deaths), x = covid_deaths, fill = state)
) +
  geom_col() +
  geom_text(
    aes(label = comma(covid_deaths)),
    hjust = 1.1,
    color = "white",
    fontface = "bold",
    size = 3.5
  ) +
  scale_fill_hue() +
  scale_x_continuous(
    breaks = seq(0, max(covid_top10$covid_deaths, na.rm = TRUE), by = 15000),
    labels = comma,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Top 10 States by COVID-19 Deaths",
    x = "COVID-19 Deaths",
    y = "State"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    legend.position = "none"
  )
```

The `Top 10 States by COVID-19 Deaths` plot shows that California has the highest number of COVID-19 deaths, followed by Texas and Florida, similar trend to total deaths. It's a good idea to perhaps explore how age groups and race/origin groups are affected, or effect these trends as well.

```{r}
fit1 <- lm(covid_deaths ~ age_group, data = cdc_clean)
summary(fit1)

fit2 <- lm(covid_deaths ~ race_origin, data = cdc_clean)
summary(fit2)
```

`Fit1` shows that there is a statistically significant relationship between `age group` and `COVID-19 deaths` (p < 2.2e-16), with older age groups having higher COVID-19 death counts. The model explains approximately 85% of the variation in COVID-19 deaths among different age groups (R2 = 0.8505; adjusted R2 = 0.8478). The prediction error rate is 1,234 deaths (F= 305.3, p < 2.2e-16).

`Fit2` indicates that there is a statistically significant relationship between `race and Hispanic origin group` and `COVID-19 deaths` (p < 2.2e-16). The model explains approximately 60% of the variation in COVID-19 deaths among different race/origin groups (adjusted R2 = 0.5958). 

The prediction error rate is 1,567 deaths (F= 120.5, p < 2.2e-16). This suggests that race/origin group is an important factor in understanding COVID-19 death counts.

```{r}
library(ggplot2)

ggplot(cdc_clean, aes(x = covid_deaths)) +
  geom_histogram(bins = 40, color = "white") +
  labs(title = "Distribution of COVID-19 Deaths", x = "COVID-19 Deaths", y = "Count") +
  theme_minimal()
```

```{r}
library(dplyr)
library(ggplot2)
library(scales)

# summarize covid deaths by age group and race/origin group
p_age <- cdc_clean %>%
  mutate(group = age_group, variable = "Age Group") %>%
  group_by(variable, group) %>%
  summarise(covid_deaths = sum(covid_deaths, na.rm = TRUE), .groups = "drop")

p_race <- cdc_clean %>%
  mutate(group = race_origin, variable = "Race/Origin Group") %>%
  group_by(variable, group) %>%
  summarise(covid_deaths = sum(covid_deaths, na.rm = TRUE), .groups = "drop")

plot_dat <- bind_rows(p_age, p_race)
ggplot(plot_dat, aes(x = reorder(group, covid_deaths), y = covid_deaths, fill = group)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~variable, scales = "free_y") +
  scale_fill_hue() +
  scale_y_continuous(
    labels = scales::comma,
    breaks = scales::breaks_pretty(n = 6)
  ) +
  labs(
    title = "COVID-19 Deaths by Age Group and Race/Origin Group",
    x = "",
    y = "COVID-19 Deaths"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "none"
  )
```

This graph looks at COVID-19 deaths by age group and race/origin group. The left panel shows that COVID-19 deaths increase with age, with the highest counts in the 85+ age group. The right panel shows that certain race/origin groups, such as Non-Hispanic Black and Hispanic, have higher COVID-19 death counts compared to Non-Hispanic White and Non-Hispanic Asian groups. While it shows disparity in deaths across race/origin groups, it is important to consider other factors such as socioeconomic status and access to healthcare, which are not shown in this dataset.

Note: Chatgpt was used to assist with code writing, color palettes, and brevity. All interpretations and conclusions were made by me.

### Citation
National Center for Health Statistics. Provisional COVID-19 Deaths by Race and Hispanic Origin, and Age. Date accessed [11 Februrary 2026]. Available from https://data.cdc.gov/d/ks3g-spdg.